Assistant Tools and Accessibility Features for Blind People Playing Visual-Centric Digital Games

### 2024-02-18 Paper/paper.tex
Fully implemented section "The Details" and "Conclusions and Further Work"

	So, the main idea of this paper is to give a broad insight of \emph{Universally Accessible Game Design} (UAGD) which addresses the previously mentions developing problems.
	Additionally, this study explores different and innovative accessibility features and tools, including haptic feedback and its ways to improve game experience which is a major technical challenge.

	This raises following relevant research questions (RQ) which contributes to people playing visual-centric digital games:

	\begin{itemize}
		\setlength\itemsep{0.5em}
		\item RQ1: How can UAGD improve the development of games with regard to accessibility?
		\item RQ2: Which innovative accessibility features and tools can enhance the gaming experience for blind players?
		\item RQ3: In what ways can haptic feedback enhancements improve the game experience for blind players?
	\end{itemize}

	\subsection{Motivation}
	One big step forward making mainstream games more accessible for blind people was the game \emph{The Last of Us Part II} (TLOU2) \cite{playstation_last_2020, playstation_last_2020-1}. 
	According to \textcite{leite_extended_2021} the game company \emph{Naughty Dog} implemented more than 60 accessibility features and is considered as the most accessible game ever produced.
	Additionally, \textcite{dale_last_2024} described that the game can be played all the way through with audio cues and navigation aids.
	It includes preset accessibility options for common disabilities like hearing or vision impairments. 
	It also introduces accessibility menus when the game is first started, making it easier for players with disabilities to adjust settings.
	To top that, \emph{Naughty Dog} released a remastered version of TLOU2 in 2024 with a reworked \emph{Cinematic Audio Descriptions} feature \cite{playstation_last_2024}.

	\section{The Problem} \label{sec:theProblem}
	Blind players encounter many different barriers when playing visual-centric digital games which often rely greatly on graphical interfaces and visual cues. 
	To top that, the collection of those mainstream games have different perspectives such as top-down, first-person, and third-person views, where all three views give the player unique challenges in navigating game environment, understanding game objectives and interacting with in-game elements like players or objects.
	Building on that, the authors of \textcite{goncalves_my_2023} have categorized seven themes and identified unresolved barriers (see \autoref{fig:seven-themes}) which still represent a great challenge for both players and developers.
	%TODO rewrite

@@

    To address the environment and pathfinding problems, one new technology was introduced in 2018 by \textcite{andrade_echo-house_2018} to use echolocation to explore a virtual environment which could drastically improve navigation in it.
    As for perspective (camera) and interacting with the world, hardware solutions like haptic feedback or AI assistant tools could be a solution when developed and integration further.
    Whereas the haptic feedback \cite{bello_haptics_2016} of for example PS5-Controllers \cite{akyaman_anticipated_2021, chen_gamepad_2024} could indicate when players aim too high or too low while an AI tool could provide the player with enhanced audio descriptions and how to interact with the world.

    In summary, it can be said that the game industry already takes into account the most important accessibility features, yet most modern visual-centric digital games are still a major challenge for blind players.
    In the following section we will delve deeper into some of the listed features above, their problems and possible solutions, as well as how to improve the overall experience of blind people.

@@

	The study highlights that blind players often rely on audio cues to understand and navigate game environments.
	They use repetitive actions like bumping into a wall to create a mental map of the environment.
	Players also try to create landmarks by interacting with the game world, for example, by leaving enemies behind to know in which area they are at the moment.
	This approach helps to navigate in game environments but also scare of new blind players due to frustration if the player become disorientated.

@@

	\subsection{Universally Accessible Game Design}
	In recent years, the awareness of accessible game design has been growing tremendously.
	According to \emph{World Health Organization} \cite{world_health_organization_international_2004}, one out of ten persons has disabilities which often results in limitations in hearing, memory, vision or motor functions.

@@

	\subsubsection{Technical Challenges}
	Implementing UAGD presents several technical challenges including compatibility with different hardware and the scalability of accessibility features.
	By using a user-centered approach, as presented by \textcite{grammenos_unified_2007}, developers can integrate accessibility into the game development cycle and ensure that technical challenges are considered from the beginning of the project.
	The basic steps are summarized in \autoref{fig:universal-game-design-approach}, whereas an elaboration of these is provided in \textcite{grammenos_unified_2007}.

@@

	\subsection{Assistive haptic feedback}
	%TODO
	Short introduction into feedback variants and where it is used "visual perception (sight), haptics (tactile) and aural (hearing)". \cite{kuber_towards_2007}

	\subsection{Further possibilities to use haptic feedback}
	%TODO
	Combination with echolocation?

	\subsection{Ways to improve existing accessibility features and tools}
	%TODO
	AI-tools?

	\section{Conclusions and Further Work}
	%TODO
	A recap of the problem addressed and the proposed solution.
	The implications of your research for the gaming industry and the broader accessibility community.
	Any remaining challenges or unanswered questions that need to be addressed.
	Suggestions for future research directions or enhancements to your proposed idea.
	How your work contributes to advancing the state-of-the-art in accessibility technology for blind individuals playing digital games.

>>>>

	So, the main idea of this paper is to give a broad insight of \emph{Universally Accessible Game Design} (UAGD) which addresses the previously mentioned developing problems.
	Additionally, this study explores different and innovative accessibility features and tools, including haptic feedback and its ways to improve game experience which is a major technical challenge.

	This raises following relevant research questions (RQ) that contribute to blind people playing visual-centric digital games:

	\begin{itemize}
		\setlength\itemsep{0.5em}
		\item RQ1: How can UAGD improve the development of games with regard to accessibility? (\autoref{subsec:uagd})
		\item RQ2: In what ways can haptic feedback enhancements improve the game experience for blind players? (\autoref{subsec:assistiveHapticFeedback})
		\item RQ3: Which innovative accessibility features and tools can enhance the gaming experience for blind players? (\autoref{subsec:improvements})
	\end{itemize}

	\subsection{Motivation}
	One big step forward making mainstream games more accessible for blind people was the game \emph{The Last of Us Part II} (TLOU2) \cite{playstation_last_2020, playstation_last_2020-1}. 
	According to \textcite{leite_extended_2021} the game company \emph{Naughty Dog} implemented more than 60 accessibility features and is considered as the most accessible game ever produced (See \autoref{fig:tlou-accessibility-presets}).

	\begin{figure*}[ht]
		\centering
		\includegraphics[scale=0.6,width=\textwidth]{assets/tlou-accessibility-presets.png}
		\caption{Accessibility presets in TLOU2 (Source: \textcite{playstation_last_2020-1})}
		\label{fig:tlou-accessibility-presets}
	\end{figure*}

	Additionally, \textcite{dale_last_2024} described that the game can be played all the way through with audio cues and navigation aids.
	It includes preset accessibility options for common disabilities like hearing or vision impairments. 
	It also introduces accessibility menus when the game is first started, making it easier for players with disabilities to adjust settings.
	To top that, \emph{Naughty Dog} released a remastered version of TLOU2 in 2024 with a reworked \emph{Cinematic Audio Descriptions} feature \cite{playstation_last_2024}.

	In the past, video games were not designed with accessibility mind, which resulted in bad gaming experience for people with disabilities. 
	However, over time, game companies started to develop and include accessibility features in their games which significantly improved the game experience for blind players.

	There are still many games which were not developed in a way that everyone can play it.
	This paper aims to help to find a way to help the development of games as well as exploring existing accessibly features and ways to enhance them. 
	By looking at what TLOU2 did and thinking about what else could be done, we can work towards making games a better experience for everyone.

	\section{The Problem} \label{sec:theProblem}
	Blind players encounter many different barriers when playing visual-centric digital games which often rely mainly on graphical interfaces and visual cues. 
	Additionally, games have different perspectives such as top-down, first-person, and third-person views, where all three views give the player unique challenges in navigating within game environments, understanding game objectives and interacting with in-game elements like players or objects.
	Building on that, the authors of \textcite{goncalves_my_2023} have categorized seven themes and identified unresolved barriers (see \autoref{fig:seven-themes}) which still represent a great challenge for both players and developers.
	%TODO rewrite

@@

    To address the environment and pathfinding problems, one new technology was introduced in 2018 by \textcite{andrade_echo-house_2018} to use echolocation to explore a virtual environment which could drastically improve navigation in it.
    As for perspective (camera) and interacting with the world, hardware solutions like haptic feedback or \emph{Artificial Intelligence} (AI) tools could be a solution when developed and integrated further.
    Whereas the haptic feedback \cite{bello_haptics_2016} of for example PS5-Controllers \cite{akyaman_anticipated_2021, chen_gamepad_2024} could indicate the environment they are walking on by adjusting the vibrations.

    The goal is to find an answer to the research question above by researching the field of UAGD, how to use haptic feedback to improve accessibility features and to find new ways to enhance existing accessibility features and tool in combination with AI.
    Therefore, game developers should use the UAGD as a guide and ensure that all important accessibility features listed above are included in their games.

    The following section provides a deeper insight into related work that has contributed to the understanding of the importance of accessibility in visual-centric digital games.

@@

	The study highlights that blind players often rely on audio cues to understand and navigate within game environments.
	They use repetitive actions like bumping into a wall to create a mental map of the environment.
	Players also try to create landmarks by interacting with the game world, for example, by leaving enemies behind to know in which area they are at the moment.
	This approach helps to navigate within game environments but also scare of new blind players due to frustration if the player become disorientated.

@@

	\subsection{Universally Accessible Game Design} \label{subsec:uagd}
	In recent years, the awareness of accessible game design has been growing tremendously.
	According to \emph{World Health Organization} \cite{world_health_organization_international_2004}, one out of ten persons has disabilities which often results in limitations in hearing, memory, vision or motor functions.

@@

	\subsubsection{Technical Challenges}
	Implementing UAGD presents several technical challenges including compatibility with different hardware and the scalability of accessibility features.
	By using a user-centered approach, as presented by \textcite{grammenos_unified_2007}, developers can integrate accessibility into the game development cycle and ensure that technical challenges are considered from the beginning of the project.
	The basic steps are (See \autoref{fig:universal-game-design-approach}):

	\begin{itemize}
		\setlength\itemsep{0.5em}
		\item Abstract task-based game design
		\item Polymorphic specialization with design alternatives
		\item Appropriateness analysis for the design alternatives
		\item Compatibility analysis among design alternatives
		\item Prototyping
		\item Usability and accessibility evaluation
	\end{itemize}

	An elaboration of these steps is provided in \textcite{grammenos_unified_2007}

@@

	\subsection{Assistive haptic feedback} \label{subsec:assistiveHapticFeedback}
	Assistive haptic feedback gets increasingly more important, especially in the gaming industry \cite{kuber_towards_2007}.
	In this section we delve into innovative approaches to enhance the gaming experience for blind player through tactile technologies.
	First we will give an overview of sensory perception which consists of visual (sight), haptic (tactile), and aural (hearing) as shown in \autoref{fig:sensory-perception}.
	Then we will explore different possibilities to enhance game experience with haptic feedback.

	\begin{figure}[ht]
		\centering
		\includegraphics[scale=0.4]{assets/sensory-perception.png}
		\caption{Sensory perception (Source: \textcite{sanfilippo_perspective_2022})}
		\label{fig:sensory-perception}
	\end{figure}

	\subsubsection{Sensory Perception in Gaming for Blind Players}
	Gaming experience for blind people primarily rely on non-visual sensory to navigated and interact within the game environment.
	Visual perception which the player mainly get though sight, is often the primary interaction in most games.
	However, for blind player, haptic and aural senses are the key.,

	Haptic feedback is the sense of touch providing a physical response by vibration to the player \cite{chen_gamepad_2024, kuber_towards_2007, bello_haptics_2016}, which represent various action in a game like drawing a bow or shooting a weapon.
	For example, a haptic glove could simulate the sensation of raindrops falling on a player's hand or the resistance encountered when pushing against an in-game object.

	Aural feedback, in the other hand, provides sound to transfer information.
	For example, the sound of footsteps can indicate that an ally or enemy is approaching, while footstep sounds changes indicate a transition from one area to another, such as moving from a city street into a quiet building.

	\subsubsection{Enhancing Gaming Experience through Assistive Haptic Feedback}
	Assistive haptic feedback helps blind players feel and interact with game worlds through touch.
	This form of feedback offers a new dimension of immersion and accessibility.
	For example, vibration can be designed in a way that they give information to the player by providing them with different vibration patterns on various surfaces such as uneven roads or a sandy beach.
	By doing that, players receive a sense of location and movement within the game environment.

	Beyond that, advanced haptic technologies like shape-changing interfaces or dynamic touch surfaces \cite{rasmussen_shape-changing_2012} can simulate more complex interactions.
	These could include the sensation of touching sand or feel the resistance when drawing a bowstring by using a haptic glove mentioned previously.
	This feedback technology could also help in navigating within a game environment where pulses could indicate the direction a player has to go.
	This not only makes games more accessible to blind players but also increase the dimension of immersion for all players.

	Considering the impact of standardizing this technology as a guideline within UAGD, making it a straightforward implementation for developers would help the game industry tremendously in terms of accessibility.

	\subsection{Ways to improve existing accessibility features and tools} \label{subsec:improvements}
	In this section we explore strategies to improve accessibility for blind players.
	We are focusing us here on two promising advanced technology ideas:

	\begin{itemize}
		\setlength\itemsep{0.5em}
		\item AI-Enhanced Testing and Audio Descriptions
		\item Combining AI with echolocation techniques
	\end{itemize}

	\subsubsection{AI-Enhanced Testing and Audio Descriptions}
	By implementing AI into the cycle of game development, we could significantly improve the accessibility for blind players by integration AI into the testing phase of the game.
	After collecting enough data about accessibility, the AI could automate the testing phase, identifying accessibility issues or UAGD guideline violations which are not immediately visible for the developer.
	This would ensure that games are designed the right way from the start.

	Additionally, by letting the AI collect data during game play tests, it could generate real-time audio descriptions of game environments and action which could improve the gaming experience for blind players.
	The generated descriptions can then customize for the needs of individual players.

	To implement this method, additional research and testing are required.
	This approach would greatly help developers in designing future games.

	\subsubsection{Combining AI with Echolocation for Navigation}
	Combining AI with echolocation technology could assist blind players in navigating within game environments.
	The AI could interpret the echolocation signals which could be combined with haptic or aural feedback, helping players understand their environment.
	For example, a glove or body armor as haptic feedback devices could receive the signals, process them to create a detailed map of the surroundings which provides the player with a much better understanding what is happening.

	For this technology to be fully realized, further research like \textcite{andrade_echo-house_2018} is required.
	This approach is promising for not only enhancing the experience for blind but all players.

	\section{Conclusions and Further Work}
	This paper addressed the challenge of enhancing accessibility features in digital games for blind players.
	Our proposed solution is centered around \emph{Universally Accessible Game Design} (UAGD) and innovative assistive technologies like haptic feedback and echolocation.
	This solution aims to improve the needs of blind players in current games.

	\subsubsection{Conclusions}
	Our findings offer answers to the proposed research questions, highlighting how UAGD (RQ1) can improve game development for better accessibility, how advancements in haptic feedback (RQ2) can significantly enhance the gaming experience not only for blind but all players and ways to improve existing accessibility features and tools (RQ3) by using AI.

	\subsubsection{Further Work}
	There are still open challenges such as the full integration of echolocation techniques as well as additional research in using AI in testing phases and audio description generation.
	Future studies should focus on the echolocation technology for gaming, exploring AI in creating dynamic audio descriptions and identifying missing accessibility features, and developing more haptic feedback systems.

	Our study opens the door to new possibilities in making digital gaming and development a better experience, ensuring that advancements in gaming technology can be enjoyed by everyone, including those who are blind.
